---
title: 'DV_RProject3 : Electricity Rates For Investor Owned and Non-Investor Owned
  Utilities'
author: "Syairah Ridzuan, Madeline Packard, Arthur DeVitalis"
date: "March 4, 2015"
output: html_document
---

Investor Owned Utilities (IOU) is defined as utilities that are owned by private companies instead of state or federal public agency (EnergyVortex.com). Both datasets used in this projects are made public by the National Renewable Energy Laboratory (NREL) and Ventyx in February 2011. The datasets contain the average electricity rates for residential, commercial and industrial purposes for each utility.

##Loading the Datasets from Oracle Database Into R Studio

The datasets pulled from the Oracle database include the average electricity rates for residential, commercial and industrial purposes for each utility, details about the companies that provide the utilities as well as the zipcodes where the utilities are located. 

```{r, include=FALSE, results = "hide"}
source("../01 Data/packages.R", echo = FALSE)
```

```{r}
source("../01 Data/io_df.R", echo = TRUE)
```

We realized that the datasets lack one valuable information which is a column that contains the states where the utilities are located. Nonetheless, with the information that we have on zipcodes, we are able to create the 'state' column and identify the states based on the corresponding zipcodes. Thus, we added 'state' column on both datasets.

```{r}
source("../01 Data/adding_io_state.R", echo = TRUE)
```

```{r}
source("../01 Data/non_io_df.R", echo = TRUE)
```

```{r, include=FALSE}
source("../01 Data/adding_non_io_state.R", echo = FALSE)
nonio_df %>% tbl_df
```

Now, both `io_df` and `nonio_df` have seven variables in their datasets. To explore the datasets, we load the datasets from Oracle but set the `returnDimensions = 'True'`. This is done so that we can recognize how many qualitative and quantitative variables present in the datasets. Then, we proceed to plot all the variables againsts the count of their own records so that we are able view the distribution of records for each of the variable. 

The following `myplot` function entails the plotting of categorical variable in the form of point plots while `myplot1` function plots the quantitative variables in the form of histograms.

```{r, message=FALSE}
source("../01 Data/plot_function_qualitative.R", echo = TRUE)
```

Notice `theme(axis.text.x=element_blank())` inside the code above; this removes the labels on the x-axes for all qualitative variables. This is because there are a lot of distinct datapoints in both datasets and they eventually made the axes become unreadable. In addition, company name and company id give us less insights compared to the number of utilities they manage which could possibly be located across many states.

```{r}
source("../01 Data/categoricals_io.R", echo = TRUE)
```

```{r, message=FALSE}
source("../03 Visualizations/io_qualitative.R", echo = TRUE)
```

```{r, message=FALSE, out.width = '1000px', out.height = '700px'}
source("../03 Visualizations/storing_io_qualitative.R", echo = FALSE)
```

```{r, message=FALSE}
source("../01 Data/plot_function_quantitative.R", echo = TRUE)
```

```{r, message=FALSE}
source("../03 Visualizations/io_quantitative.R", echo = TRUE)
```

```{r, message=FALSE, out.width = '1000px', out.height = '700px'}
source("../03 Visualizations/storing_io_quantitative.R", echo = FALSE)
```

The same steps above are applied to non-investor owned utilities dataset. The following shows the visualizations produced using `myplot` and `myplot1` functions on 'nonio_df' dataset.

```{r, include=FALSE}
source("../01 Data/categoricals_non_io.R", echo = TRUE)
```

```{r, message=FALSE, include=FALSE}
source("../03 Visualizations/non_io_qualitative.R", echo = TRUE)
```

```{r, message=FALSE, out.width = '1000px', out.height = '1500px'}
source("../03 Visualizations/storing_non_io_qualitative.R", echo = TRUE)
```

```{r, message=FALSE, include=FALSE}
source("../03 Visualizations/non_io_quantitative.R", echo = TRUE)
```

```{r, message=FALSE, out.width = '900px', out.height = '700px'}
source("../03 Visualizations/storing_non_io_quantitative.R", echo = TRUE)
```

##Joining Datasets

Using the `dplyr` package, we can combine both datasets and yield quite valuable information.

For instance, the following code combines both datasets using inner join, which matches records based on their zipcode areas. It is important to note that we avoid using `"COMPANY NAME"` and `"COMPANY ID"` for comparing the two datasets. This is because a utility company can either be a private enterprise or a government-run agency but it could not be both of the entities.

```{r}
source("../01 Data/inner_join.R", echo = TRUE)
```

The result of an inner join gives us the areas that have both investor owned (IO) and non-investor owned utilities. Notice in the table that there are some duplicates in either IO utilities on non-IO utilities - this may indicate that those areas have either more than one IO or non-IO utilities.


The following code demonstrates the full outer join operation done on both datasets.

```{r}
source("../01 Data/full_outer_join.R", echo = TRUE)
```

The full outer join operation simply gives all the IOUs and non-IOUs for all the zipcodes available in both datasets. This is very useful for viewing a complete dataset on electricity rate regardless of the provider. However, in this project, we are more interested to compare the electricity rates between the two types of provider.

Anti join is an operation that returns the records that only exists in one dataset but not in the other. For instance, the anti join below gives us areas that only have investor owned utilities (IOU). 

```{r}
source("../01 Data/unique_io.R", echo = TRUE)
```

Meanwhile, the following anti join gives us areas where electricity are only provided by non-investor owned utilities (non-IOU).

```{r}
source("../01 Data/unique_non_io.R", echo = TRUE)
```


##Visualizations From Wrangled Datasets

The join operations gave us some interesting visualization ideas. We begin by comparing the rates for areas that are only supplied by either IOUs or non-IOUs. To do this, we have to use the results from the anti joins that we did earlier, which yield `unique_io` and `unique_nonio`. We chose to visualize them using violin plot.

```{r, warning=FALSE, out.width = '900px', out.height = '500px'}
source("../02 Data Wrangling/areas with_only_io_or_nonio.R", echo = TRUE)
```

Based on the plot above, it is interesting to note that areas with only investor owned utilities have wider distribution, as suggested by the length of the red plot, compared to areas with only non-investor owned utilities which are more condensed, as depicted by the larger vertical width of the blue plot, and this comparison seems to hold for all type of rates. However, it is curious to see that distributions of rates for non-investor owned utilities are very skewed 
to the right of the plot and you will later recognize the reason behind this as we get to the later visualizations.

We pivot our focus to areas that have both IOUs and non-IOUs. Again, we use violin plot to visualize so that it will be much easier to compare with the previous plot.

```{r, warning=FALSE, out.width = '900px', out.height = '500px' }
source("../02 Data Wrangling/areas_with_both_io_and_nonio.R", echo = TRUE)
```

There are a few interesting insights in the plot above: firstly, for non-investor owned utilities, notice that the distributions for all the rates are very similar. Comparing the distribution of rates between IOU and non-IOU, it can be seen that IOU rates are less varied that the non-IOU rates. What's more surprising to us is that IOUs have much lower industrial rate than the non-IOUs; the interquartile range of IOU's industrial rate is below the first 50% of the non-IOU's industrial rate distribution. 

From the first two plots, it is evident that electricity rates for the investor owned utilities (IOU) tend to vary less (in terms of distribution) in areas where both types of utilities exist. In fact, in areas where both IOUs and non-IOUs exist, the electricity rates for IOUs tend to be lower then those provided by non-IOUs.

In the previous two violin plots, we can observe that there are many outliers in the distribution. We seek to discover which areas have significantly higher electricity rates. However, since there are many zipcodes, we decided to use state instead so that we are able to intepret the graph more effectively. Thus, we plot the distribution rates againsts their respective states according to the type of rate for investor owned utilities.

```{r, warning=FALSE, out.width = '900px', out.height = '500px'}
source("../02 Data Wrangling/distribution_rates_by_state.R", echo = TRUE)
```

Based on the plot above, it can be seen that Hawaii have significantly higher electricity rates for all types of rate compared to other states. It is also interesting to note here that the distribution between the three types of rate in within a state are quite similar for all states. However, some states like Massachussetts, Arizona and Maine have significantly wider distribution of electricity costs across all rates compared to other states.

The graph below plots the distribution rates of states according to the type of rate for non-investor owned utilities.

```{r, warning=FALSE, out.width = '900px', out.height = '700px'}
source("../02 Data Wrangling/distribution_nonio_rates_by_state.R", echo = TRUE)
```

Comparing the graph above with the previous graph on IOUs, it can be seen that non-IOUs provided electricity at much lower costs for all three types of rates, compared to IOUs, in almost all states, with the exception of Alaska - there are no significantly wide distribution of electricity rates compared to IOUs' point plot previously. Also, looking at Alaska and glancing at the scale of 'Value ($)' on the horizontal axis seems to tell us that it is Alaska's electricity rate that may have skewed the distribution rate of non-IOUs on the first violin plot above.

Note that state-by-state comparison between IOUs and non-IOUs point plots is not applicable to states; Delaware, for instance, does not have non-IOUs and Armed Forces Pacific does not have IOUs.

The following bar graphs seek to display IOUs' average electricty costs for all types of rate for each state.

```{r, warning=FALSE, out.width = '900px', out.height = '700px' }
source("../02 Data Wrangling/mean_rate_state_io.R", echo = TRUE)
```

It can be observed from the graph above that Hawaii has the highest mean cost for all types of electricity. Accessibility issue may be the cause of such expensive electricity rate. Setting Hawaii aside, California, Delaware and Florida have the highest electricity cost for residential areas. Meanwhile, California, Delaware and Washington D.C. have the most expensive electrity cost for commercial purpose. For industrial purposes, California, Florida and Massachusetts have the highest electricity rates compared to all the states.

The following graph shows non-IOUs' average electricty costs for all types of rate for each state.

```{r, warning=FALSE, out.width = '900px', out.height = '700px' }
source("../02 Data Wrangling/mean_rate_state_non_io.R", echo = TRUE)
```

Alaska have significantly higher electricity cost for commercial and residential purposes. We decided to treat Guam and Armed Forces Pacific as special cases because there may be many overhead costs that needed to be covered in order to provide electricity in these places. So setting them aside, we can see that Washington D.C. has the highest cost for commercial and residential purposes while California has the most expensive electricity rate for industrial purposes.
